---
title: "SPECIES web framework API functions"
author: "Enrique Del Callejo-Canal"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{get_species_}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
```

[SPECIES web platform](http://species.conabio.gob.mx) is an interactive tool for
the analysis of ecological niches and forecast species potential distribution,
as well as to build Complex Inference Networks to identify potential species interactions.
SPECIES development is supported by [C3-Centro de Ciencias
de la Complejidad](http://www.c3.unam.mx/), Universidad Nacional Autónoma de México and the
Comisión Nacional para el Conocimiento y Uso de la Biodiversidad ([CONABIO](https://www.gob.mx/conabio)).

This tutorial is a basic intriduction to use the r functions that connects to SPECIES api. This functions are contained in the developing package `rspecies`. `rspecies` is allocated in github public repositorie https://github.com/edelcallejoc/rspecies.  

## Vignette Info

The information contained in this vignette is corresponds to the usages of r functions to interact with SPECIES api. There are four functions that allows to extract information about SPECIES database's names. Each function can be used for many porpuses from consulting species names to adjust models in several ways.


## Consulting taxon names from SPECIES framework (get_species_names)

The simplest way to use **get_species_names** is for cheking species names available for a specific taxon rank. For example, the names of the species of `Aedes` genus.

```{r getnames1, message=FALSE, results = 'markup'}
library(rspecies)

aedes <- get_species_names(level = "genus", name = "Aedes", id = FALSE)

head(aedes, n=5)

```

A more complex example is to extract species' name from a higher taxon level. For example, The names of the species of `Mammalia` class.

```{r getnames2, message=FALSE, results = 'markup'}
mammal <- get_species_names(level = "class", name = "Mammalia", sublevel = "specie", id = FALSE)

head(mammal, n=5)

```

It is important to declared `id = FALSE` in order to avoid extensive computation. The `id` argument is intended to be used for individual queries by species. 

## Extracting species' coordinates from SPECIES framework (get_species_coords)

The function **get_species_coords** is made to extract the geographical coordinates of observations by species from the API of the SPECIES framework. The simplest way to use this function is to indicate the `species` argument with the name of a particular species, eg _Lynx rufus_.

```{r getcoords, message=FALSE, results = 'markup'}
LR_coords <- get_species_coords(species = "Lynx rufus")

head(LR_coords, n=5)

```

The `date` argument is used to add collection date information, by default this argument is defined as TRUE. While the argument `from` is still in development and at this moment it has no use.

```{r getcoords2, message=FALSE, results = 'markup'}
LR_coords_ndate <- get_species_coords(species = "Lynx rufus", date = FALSE)

head(LR_coords_ndate, n=5)

```

## Downloading the map grid from the SPECIES framework (get_species_grid)

The download of the grid is done with the function **get_species_grid**, which takes as argument the resolution of the grid, (`grid_res`). There are 4 possible values for the argument, 8,16,32 and 64.

```{r getgrid, message=FALSE, results = 'markup', fig.align='left', fig.width=5, fig.show='asis'}
Mex_grd32 <- get_species_grid(grid_res = 32)

# using ggplot
library(ggplot2)

ggplot(fortify(Mex_grd32), aes(x = long, y = lat, group = id)) +
  geom_polygon() +
  geom_path(color="white") +
  coord_map()

```

## Download model results from the SPECIES platform. (get_species_georel)

The **get_species_georel** function is used to download the results of the model set in the SPECIES platform. For the moment the function is useful to extract the results of the model for a single species. However, it is possible to use the `apply` family of functions to extract results from different models (do this operation with care since it can take a lot of time).


```{r getgeorel, message=FALSE, results = 'markup'}

# Lynx rufus model with all Mammals
LR_Model <- get_species_georel(target = "Lynx rufus",
                               start_level = "class",
                               value = "Mammalia",
                               validation = FALSE,
                               fossil = FALSE,
                               no_date = TRUE,
                               min_occ = 5,
                               bioclim = FALSE)

head(LR_Model, n = 5)
  
```